---
title: "EDA_PS3_GROUP13"
author: "Ayesha Bhimdiwala, Pulkit Maloo and Vaishnavi"
date: "1/29/2018"
output: pdf_document
---

```{r}
library(ggplot2)
library(NHANES)
library(broom)
```

#INTRODUCTION
###High systolic blood pressure is a strong predictor of heart attacks and strokes. We are trying to find a relationship of average systolic blood pressure with age, height, and weight respectively using the NHANES data which consists of the following variables:
#### BPSysAve (the average of three measurements of systolic blood pressure)
#### Age (in years; 80 or older is recorded as 80)
#### Weight (in kilograms)
#### Height (in centimeters)
#### Gender (male or female)

```{r}
#View(NHANES)
attach(NHANES)
nhanes.data = data.frame(BPSysAve , Age, Height, Weight, Gender)
colSums(is.na(nhanes.data))
n = nrow(nhanes.data)
```

We have 1449 missing values of BPSysAve, 353 missing values of Height and 78 missing values of Weight.
Since our dataset is large, we can ignore them.

Let's try to find the relationship of average systolic blood pressure with age first

#Section 01 - Relationship between Age vs Average Systolic Blood Pressure
```{r echo=FALSE}

males = subset(nhanes.data, Gender=="male")
females = subset(nhanes.data, Gender=="female")

detach(NHANES)
attach(nhanes.data)
ggplot(nhanes.data, aes(x = Age+runif(n, -0.5, 0.5), y = BPSysAve+runif(n, -0.5, 0.5), colour = Gender)) +
  geom_point(size = 0.5, alpha = 0.7) + geom_smooth(method = "loess", method.args = list(degree = 1, family="symmetric" )) + 
  labs(x = 'Age  (in years)', y = 'Systolic Blood Pressure', title = 'Average Systolic Blood Pressure by Age')

#males vs residuals to check for trend
detach(nhanes.data)
attach(males)
model_age.mal.lm = loess(BPSysAve ~ Age, data = males, degree=1)
model_age.mal.lm.df = augment(model_age.mal.lm)

ggplot(model_age.mal.lm.df, aes(x = Age+runif(nrow(model_age.mal.lm.df), -0.5, 0.5), y = .resid+runif(nrow(model_age.mal.lm.df), -0.5, 0.5))) + 
  geom_point(size = 0.5, alpha = 0.7,color="blue") + 
  geom_smooth(method = "loess", method.args = list(degree = 2, family="symmetric" )) + 
  geom_abline(slope = 0, intercept = 0) +
    labs(x = 'Age (in years)', y = 'Residuals', title = 'Residual Plot for Males')
var(model_age.mal.lm.df$.fitted)/var(model_age.mal.lm.df$BPSysAve)

detach(males)
#females vs residuals to check for trend
attach(females)
model_age.fem.lm = loess(BPSysAve ~ Age, data = females, degree=1)
model_age.lm.fem.df = augment(model_age.fem.lm)

ggplot(model_age.lm.fem.df, aes(x = Age+runif(nrow(model_age.lm.fem.df), -0.5, 0.5), y = .resid+runif(nrow(model_age.lm.fem.df), -0.5, 0.5))) + geom_point(size = 0.5, alpha = 0.7, color="red") +
  geom_smooth(method = "loess", method.args = list(degree = 2, family="symmetric" ), color = "red") + 
  geom_abline(slope = 0, intercept = 0) +
    labs(x = 'Age (in years)', y = 'Residuals', title = 'Residual Plot for Females')
var(model_age.lm.fem.df$.fitted)/var(model_age.lm.fem.df$BPSysAve)

detach(females)
```
Both the male and female heights are right-skewed with 353 missing values. Since, the data is right-skewed we took median of the heights to handle missing values. The data seems to be rounded, so we unround it by adding random noise.

#Section 02 - Relationship between Height vs Average Systolic Blood Pressure

```{r echo=FALSE}

attach(nhanes.data)
ggplot(nhanes.data, aes(x = Height+runif(n, -0.5,0.5), y = BPSysAve+runif(n, -0.5,0.5), colour = Gender)) +
  geom_point(size = 0.5, alpha = 0.7) + geom_smooth(method = "loess", method.args = list(degree = 1, family="symmetric")) + 
  labs(x = 'Height (in cms)', y = 'Systolic Blood Pressure', title = 'Average Systolic Blood Pressure by Height')

#males vs residuals to check for trend
detach(nhanes.data)
attach(males)
model_Height.lm.mal = loess(BPSysAve ~ Height, data = males, degree=1)
model_Height.lm.mal.df = augment(model_Height.lm.mal)

ggplot(model_Height.lm.mal.df, aes(x = Height+runif(nrow(model_Height.lm.mal.df),-0.5,0.5), y = .resid+runif(nrow(model_Height.lm.mal.df),-0.5,0.5))) + 
  geom_point(size = 0.5, alpha = 0.7,color="blue") + 
  geom_smooth(method = "loess", method.args = list(degree = 2, family="symmetric" )) + 
  geom_abline(slope = 0, intercept = 0) +
    labs(x = 'Height (in cms)', y = 'Residuals', title = 'Residual Plot for Males')
var(model_Height.lm.mal.df$.fitted)/var(model_Height.lm.mal.df$BPSysAve)
detach(males)
#females vs residuals to check for trend
attach(females)
model_Height.lm.fem = loess(BPSysAve ~ Height, data = females, degree=1)
model_Height.lm.fem.df = augment(model_Height.lm.fem)

ggplot(model_Height.lm.fem.df, aes(x = Height+runif(nrow(model_Height.lm.fem.df),-0.5,0.5), y = .resid+runif(nrow(model_Height.lm.fem.df),-0.5,0.5))) + geom_point(size = 0.5, alpha = 0.7, color="red") +
  geom_smooth(method = "loess", method.args = list(degree = 2, family="symmetric" ), color = "red") + 
  geom_abline(slope = 0, intercept = 0) +
    labs(x = 'Height (in cms)', y = 'Residuals', title = 'Residual Plot for Females')
var(model_Height.lm.fem.df$.fitted)/var(model_Height.lm.fem.df$BPSysAve)
detach(females)

```
Both the male and female weights are left-skewed with 78 missing values. Since, the data is right-skewed we took median of the weights to handle missing values. 
7% variation for males


#Section 03 - Relationship between Weight vs Average Systolic Blood Pressure

```{r echo=FALSE}

attach(nhanes.data)
ggplot(nhanes.data, aes(x = Weight+runif(n, -0.5,0.5), y = BPSysAve+runif(n, -0.5,0.5), colour = Gender)) +
  geom_point(size = 0.5, alpha = 0.7) + geom_smooth(method = "loess", method.args = list(degree = 1, family="symmetric" )) + 
  labs(x = 'Weight (in kgs)', y = 'Systolic Blood Pressure', title = 'Average Systolic Blood Pressure by Weight')

#males vs residuals to check for trend
detach(nhanes.data)
attach(males)
model_Weight.mal.lm = loess(BPSysAve ~ Weight, data = males, degree=1)
model_Weight.lm.mal.df = augment(model_Weight.mal.lm)

ggplot(model_Weight.lm.mal.df, aes(x = Weight+runif(nrow(model_Weight.lm.mal.df),-0.5,0.5), y = .resid+runif(nrow(model_Weight.lm.mal.df),-0.5,0.5))) + 
  geom_point(size = 0.5, alpha = 0.7,color="blue") + 
  geom_smooth(method = "loess", method.args = list(degree = 2, family="symmetric" )) + 
  geom_abline(slope = 0, intercept = 0) +
    labs(x = 'Weight (in kgs)', y = 'Residuals', title = 'Residual Plot for Males')
var(model_Weight.lm.mal.df$.fitted)/var(model_Weight.lm.mal.df$BPSysAve)

detach(males)
#females vs residuals to check for trend
attach(females)
model_Weight.fem.lm = loess(BPSysAve ~ Weight, data = females, degree=1)
model_Weight.fem.lm.df = augment(model_Weight.fem.lm)

ggplot(model_Weight.fem.lm.df, aes(x = Weight+runif(nrow(model_Weight.fem.lm.df),-0.5,0.5), y = .resid+runif(nrow(model_Weight.fem.lm.df),-0.5,0.5))) + geom_point(size = 0.5, alpha = 0.7, color="red") +
  geom_smooth(method = "loess", method.args = list(degree = 2, family="symmetric" ), color = "red") + 
  geom_abline(slope = 0, intercept = 0) +
    labs(x = 'Weight (in kgs)', y = 'Residuals', title = 'Residual Plot for Females')
var(model_Weight.fem.lm.df$.fitted)/var(model_Weight.fem.lm.df$BPSysAve)
detach(females)
```

#CONCLUSION